### 2017-01-17 text/Chapter05.tex
Edits pages 95-124

\item For two sample coeffect calculi discussed earlier (dataflow and implicit parameters),
  we give reduction rules for the comonadically-inspired primitives and we extend the progress
  and preservation proofs, showing that well-typed programs produced by translation from two
  coeffect languages do not go wrong (Section~\ref{sec:semantics-proofs})

@@

The key claim of this thesis is that writing context-aware programs using coeffects is easier and
less error-prone. In this chapter, we substantiate the claim by showing that programs written in the
coeffect calculus and evaluated using the translation provided here do not ``go wrong''.

To provide an intuition, consider two context-aware programs. The first calls a function that
adds two implicit parameters in a context where one of them is defined. The second calculates
the difference between the current and the previous value in a dataflow computation.
For comparison, we show the code written in a coeffect dataflow language (on the left) and
using standard ML-like libraries (on the right):
%
\begin{equation*}
\begin{array}{lcl}
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}`\rightarrow`\\[-0.2em]
 \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?one} = 10~\kvd{in}\\[-0.2em]
\ident{add}~0
\end{array}
&\quad&
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}~\ident{params}\rightarrow\\[-0.2em]
 \quad \ident{lookup}~\str{one}~\ident{params}~+ \\[-0.2em]
 \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
\ident{add}~0~(\ident{cons}~\str{one}~10~\ident{params})
\end{array}
\\[3.5em]
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{x}-\kvd{prev}~\ident{x}
\end{array}
&&
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
\end{array}
\end{array}
\end{equation*}
%
The \ident{add} function (on the left) has a type $\ident{int}\xrightarrow{\cclrd{ \{ \ident{?one},\ident{?two} \} }}\ident{int}$.
We call it in a context containing $\cclrd{\ident{?one}}$ and so the coeffect of the program is
$\cclrd{ \{ \ident{?two} \} }$. The safety property for implicit parameters
(Theorem~\ref{thm:semantics-sound-impl}) guarantees that, when executed in a context that
provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to
a value of the correct type (or never terminates).

If we wrote the code without coeffects (on the right), we could use a dynamic map to pass around
a dictionary of parameters (the \ident{lookup} function obtains a value and \ident{add} adds a new
assignment to the map). In that case, the type of \ident{add} is just $\ident{int}\rightarrow\ident{int}$
and so the user does not know which implicit parameters it will need.

@@

In summary, a coeffect type system, captures certain runtime
demands of context-aware programs and (as we show in this chapter), eliminates common errors
related to working with context.

@@

The correspondence between effect system and monads has been pointed out by Wadler and Thiemann
\cite{monads-effects-marriage} and further explored by Atkey \cite{monads-parameterised-notions}
and Vazou and Leijen \cite{monads-effects-remarriage}). This line of work relates effectful
functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations
$\tau_1 \rightarrow \mtyp{\sigma} \tau_2$. In this chapter, we show a similar correspondence between
\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus,
defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
of the work on effect systems and monads.

@@

For a well-defined semantics, we need to ensure that a well-typed term is assigned exactly one
meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
\cite{semantics-bltres} and show the morhphisms assigned to multiple typing derivations are equivalent.
Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
using the unique typing derivation specified in Section~\ref{sec:flat-unique}.

As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful
computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that
approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$,
for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models
failures (the Maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list
monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$
(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing
monadic computations -- sequential composition and strength for lifting free variables into the
body of computation under a lambda abstraction.

@@

The monadic and comonadic model outlined above represents at most a binary analysis of effects or
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an
over-simplification as we can use \eg~stacks of monad transformers and model functions with
two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers
require the user to define complex systems of lifting to be composable. Consequently, they are usually used
for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
stateful computation).}.

@@

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta)
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

\noindent
From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries
additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function
$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
the context-dependent function to value(s) in the context and then propagates the context.

@@

\begin{example}[Non-empty list]
A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:

@@

\begin{example}[Indexed product]
The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
We assume that, all implicit parameters have the type $\ident{num}$. The data type $\ctyp{\cclrd{r}}{\alpha}
= \alpha \times (\cclrd{r} \rightarrow \ident{num})$ represents a value $\alpha$ together with a function that
associates a parameter value $\ident{num}$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
The cobind and counit operations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
\end{array}
\quad
\begin{array}{l}
\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta)
    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
\end{array}
\end{equation*}
\end{example}

\noindent
In the definition, we use the notation $(a, g)$ for a pair containing a value of type $\alpha$
together with $g$, which is a function of type $\cclrd{r} \rightarrow \ident{num}$. The \ident{counit} operation
takes a value and a function (with empty set as a domain), ignores the function and extracts the
value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to
restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
implicit parameters required by the argument of $f$ and by the resulting computation, respectively
(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed
to \ident{cobind} is defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is
valid in both cases.

@@

To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions}
require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
%
\begin{equation*}
\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
\end{equation*}
%
The \ident{m} operation is needed in the semantics of lambda abstraction. Semantically, it represents
merging of contextual capabilities attached to the variable contexts of the declaration site (containing free variables)
and the call site (containing bound variable). For example, for implicit parameters, this combines
the additional parameters defined in the two contexts.

@@

\begin{example}[Monoidal comonads]
Just like indexed comonads generalize co\-monads, the additional structure of
flat indexed comonads generalizes the symmetric semimonoidal comonads of Uustalu
and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the
monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation
is defined by duplication.
\end{example}

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of $f$, which itself requires $m$ past values. When combining two lists, $\ident{merge}_{m,n}$
behaves as \emph{zip} and produces a list of length of the shorter argument. When splitting a list,
$\ident{split}_{m, n}$ needs the maximum of the lengths.

@@

As discussed in Section~\ref{sec:flat-unique}, different typing derivations of coeffect programs
may have different meaning (\eg~when working with implicit parameters) and so the semantics is
defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a semantics to
a term, we need to choose a particular typing derivation. The algorithm for choosing a unique
typing derivation for our three systems has been defined in Section~\ref{sec:flat-unique}.

\paragraph{Contexts and types.}
The modelling of contexts and functions generalizes the concrete examples discussed in
Chapter~\ref{ch:applications}. We use the family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed)
data-type that wraps the product of free variables of the context and the arguments of functions:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
\end{array}
\end{equation*}

@@

The semantics of application (\emph{app}) first duplicates the free-variable product inside the
context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using
$\ident{split}_{\cclrd{r}, \cclrd{s} \cpar \cclrd{t}}$. The two contexts contain the same variables
(as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
context (with index $\cclrd{r}$) is used to evaluate $e_1$ using the semantic function $f$. The
result is a function $\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. The second context
(with index $\cclrd{s}\cseq\cclrd{t}$) is used to evaluate $e_2$ and using the semantic function $g$
and wrap it with context required by the function $e_1$ by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$.
The \ident{app} operation than applies the function (first element) on the argument (second element).
Finally, numbers (\emph{num}) become constant functions that ignore the context.

@@

Although the notion of indexed comonads presented in the previous section is novel and interesting
in its own, the main reason for introducing it is that we can view it as a translation that provides
embedding of context-aware domain-specific languages in a simple target functional language.
In this section, we follow the example of effects and monads and we use the semantics to define a
translation akin to the ``do'' notation in Haskell.

A context-aware \emph{source} program written using a concrete context-aware domain-specific
language (capturing dataflow, implicit parameters or other kinds of context awareness) with
domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
is translated to a \emph{target} language that is not context-aware. The target language is
a small functional language consisting of:
%
\begin{itemize}
  \item Simple functional subset formed by lambda calculus with support for tuples and numbers.
  \item Comonadically-inspired primitives corresponding to \emph{counit}, \emph{cobind} and
    other operations of flat indexed comonads.
  \item Additional primitives that model contextual operations of each concrete coeffect language
    (\emph{prev} for the \kvd{prev} keyword, \emph{lookup} for the \ident{?p} syntax and
    \emph{letimpl} for the $\kvd{let}~\ident{?p}=\ldots$ notation).
\end{itemize}
%
The syntax, typing and reduction rules of the first part (simple functional language) are common to
all concrete coeffect domain-specific languages. The syntax and typing rules of the second part
(comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
rules} for the comonadically-inspired primitives differ -- they capture the concrete notions of
context. Finally, the third part (domain-specific primitives) will differ for each coeffect
domain-specific language.

@@

The syntax of the target programming language is shown in Figure~\ref{fig:semantics-target}.
The values include numbers $n$, tuples and function values. The expressions include variables $x$,
values, lambda abstraction and application and operations on tuples. We do not need recursion or
other data types (although a realistic programming language would include them).
In what follows, we also use the following syntactic sugar for let binding:
%
\begin{equation*}
\begin{array}{rcl}
  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2)~e_1
\end{array}
\end{equation*}
%
Finally, $K[e]$ defines the syntactic evaluation context in which sub-expressions are evaluated.
Together with the evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the
standard call-by-name semantics of the common parts of the target language. The (standard) typing rules
for the common expressions of the target language are also shown in Figure~\ref{fig:semantics-target}.

@@

\begin{lemma}[Canonical forms]
\label{thm:semantics-fp-canon}
For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
\begin{enumerate}
  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
\end{enumerate}
\end{lemma}
\begin{proof}
  For (1), the last typing rule must have been (\emph{num}); for (2), it must have been
  (\emph{abs}) and for (3), the last typing rule must have been (\emph{tup})
\end{proof}

@@

\begin{theorem}[Type preservation]
\label{thm:semantics-fp-pres}
  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow$.

@@

\begin{theorem}[Progress]
\label{thm:semantics-fp-prog}
  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
\end{theorem}
\begin{proof}
  By rule induction over $\vdash$.

@@

In this chapter, we use the same structure in a different way. Rather than treating the rules as
\emph{denotation} in categorical sense, we treat them as \emph{translation} from a source
domain-specific coeffect language into a target language with comonadically-inspired primitives
described in the previous section.

@@

\subsection{When is coeffect not a monad}
\label{sec:semantics-related-monad}

>>>>

\item For two sample coeffect calculi discussed earlier (dataflow and implicit parameters),
  we give reduction rules for the comonadically-inspired primitives and we extend the progress
  and preservation proofs, showing that well-typed programs produced by translation from two
  coeffect languages do not ``get stuck'' (Section~\ref{sec:semantics-proofs}).

@@

In Chapter~\ref{ch:intro}, we claim that coeffects make writing context-aware programs
less error-prone. In this chapter, we substantiate this claim by showing that programs written 
in the coeffect calculus and evaluated using the translation provided here do not ``get stuck''.

To provide an intuition, consider two context-aware programs. The first calls a function that
adds two implicit parameters in a context where one of them is defined. The second calculates
the difference between the current and the previous value in a dataflow computation.
For comparison, we show the code written in a coeffect dataflow language (on the left) and
using standard ML-like libraries (on the right):
%
\begin{equation*}
\begin{array}{lcl}
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?one} = 10~\kvd{in}\\[-0.2em]
\ident{add}~0
\end{array}
&\quad&
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}~\ident{params}\rightarrow\\[-0.2em]
 \quad \ident{lookup}~\str{one}~\ident{params}~+ \\[-0.2em]
 \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
\ident{add}~0~(\ident{cons}~\str{one}~10~\ident{params})
\end{array}
\\[3.5em]
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{x}-\kvd{prev}~\ident{x}
\end{array}
&&
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
\end{array}
\end{array}
\end{equation*}
%
The \ident{add} function (on the left) has a type $\ident{int}\xrightarrow{\cclrd{ \{ \ident{?one},\ident{?two} \} }}\ident{int}$.
We call it in a context containing $\cclrd{\ident{?one}}$ and so the coeffect of the program is
$\cclrd{ \{ \ident{?two} \} }$. The safety property for implicit parameters
(Theorem~\ref{thm:semantics-sound-impl}) guarantees that, when executed in a context that
provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to
a value of the correct type (or never terminates).

If we wrote the code without coeffects (on the right), we could use a dynamic map to pass around
a dictionary of parameters (the \ident{lookup} function obtains a value and \ident{add} adds a new
assignment to the map). In that case, the type of \ident{add} is just $\ident{int}\rightarrow\ident{int}$
and so there is no static account of which implicit parameters it will need.

@@

In summary, a coeffect type system captures certain runtime
demands of context-aware programs and (as we show in this chapter), eliminates common errors
related to working with context.

@@

The correspondence between effect systems and monads has been pointed out by Wadler and Thiemann
\cite{monads-effects-marriage} and further explored by Atkey \cite{monads-parameterised-notions}
and Vazou and Leijen \cite{monads-effects-remarriage}. This line of work relates effectful
functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations
$\tau_1 \rightarrow \mtyp{\sigma} \tau_2$. In this chapter, we show a similar correspondence between
\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus,
defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
of the work on effect systems and monads.

@@

For a well-defined semantics, we need to ensure that a well-typed term is assigned exactly one
meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
\cite{semantics-bltres} and show the morphhisms assigned to multiple typing derivations are equivalent.
Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
using the unique typing derivation specified in Section~\ref{sec:flat-unique}.

As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful
computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that
approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$,
for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models
failures (the Maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (set
monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$
(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing
monadic computations -- sequential composition and strength for lifting free variables into the
body of a computation under a lambda abstraction.

@@

The monadic and comonadic model outlined above represents at most a binary analysis of effects or
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an
over-simplification as we can use \eg~stacks of monad transformers and model functions with
two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers
require the user to define complex systems of lifting to be composable. Consequently, they are usually used
for capturing different classes of effects (exceptions, non-determinism, state), but not for
capturing individual effects (\eg~a set of memory regions that may be accessed by a
stateful computation).}.

@@

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a functor (mapping types to types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta)
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

\noindent
From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value $\ctyp{}{\alpha}$ 
that carries additional context. The $\ident{cobind}$ operation turns a context-dependent function
$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
the context-dependent function to value(s) in the context and then propagates the context.

@@

\begin{example}[Non-empty list]
Non-empty lists are recursive data types $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:

@@

\begin{example}[Indexed product]
The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
We assume that all implicit parameters have the type $\ident{num}$. The data type $\ctyp{\cclrd{r}}{\alpha}
= \alpha \times (\cclrd{r} \rightarrow \ident{num})$ represents a value $\alpha$ together with a function that
associates a parameter value $\ident{num}$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
The cobind and counit operations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
\end{array}
\quad
\begin{array}{l}
\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta)
    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
\end{array}
\end{equation*}
\end{example}

\noindent
In the definition, we use the notation $(a, g)$ for a pair containing a value of type $\alpha$
together with $g$, which is a function of type $\cclrd{r} \rightarrow \ident{num}$. The \ident{counit} operation
takes a value and a function (with empty set as a domain), ignores the function and extracts the
value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to
restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in order to get
implicit parameters required by the argument of $f$ and by the resulting computation, respectively
(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed
to \ident{cobind} is defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is
valid in both cases.

@@

To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions}
require a \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
%
\begin{equation*}
\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
\end{equation*}
%
The \ident{m} operation is needed in the semantics of lambda abstraction. Semantically, it represents
merging of contextual capabilities attached to the variable contexts of the declaration site (containing free variables)
and the call site (containing bound variable). For example, for implicit parameters, this combines
the additional parameters defined in the two contexts.

@@

\begin{example}[Monoidal comonads]
Just as indexed comonads generalize co\-monads, the additional structure of
flat indexed comonads generalizes the symmetric semimonoidal comonads of Uustalu
and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the
monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation
is defined by duplication.
\end{example}

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of $f$, which itself requires $m$ past values. When combining two lists, $\ident{merge}_{m,n}$
behaves as \emph{zip} and produces a list as long as the shorter argument. When splitting a list,
$\ident{split}_{m, n}$ needs the maximum of the lengths.

@@

As discussed in Section~\ref{sec:flat-unique}, different typing derivations of coeffect programs
may have different meaning (\eg~when working with implicit parameters) and so the semantics is
defined over a \emph{typing derivation} rather than over a \emph{term}. To assign a semantics to
a term, we need to choose a particular typing derivation. The algorithm for choosing a unique
typing derivation for our three systems has been defined in Section~\ref{sec:flat-unique}.

\paragraph{Contexts and types.}
The modelling of contexts and functions generalizes the concrete examples discussed in
Chapter~\ref{ch:applications}. We use the family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed)
data type that wraps the product of free variables of the context and the arguments of functions:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
\end{array}
\end{equation*}

@@

The semantics of application (\emph{app}) first duplicates the free-variable product inside the
context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using
$\ident{split}_{\cclrd{r}, \cclrd{s} \cpar \cclrd{t}}$. The two contexts contain the same variables
(as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
context (with index $\cclrd{r}$) is used to evaluate $e_1$ using the semantic function $f$. The
result is a function $\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. The second context
(with index $\cclrd{s}\cseq\cclrd{t}$) is used to evaluate $e_2$ and using the semantic function $g$
and wrap it with context required by the function $e_1$ by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$.
The \ident{app} operation then applies the function (first element) on the argument (second element).
Finally, numbers (\emph{num}) become constant functions that ignore the context.

@@

Although the notion of indexed comonads presented in the previous section is novel and interesting
on its own, the main reason for introducing it is that we can view it as a translation that provides
embedding of context-aware domain-specific languages in a simple target functional language.
In this section, we follow the example of effects and monads and we use the semantics to define a
translation akin to the ``do'' notation in Haskell.

A \emph{source} program written using a concrete context-aware domain-specific
language (capturing dataflow, implicit parameters or other kinds of context awareness) with
domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
is translated to a \emph{target} language that is not context-aware. The target language is
a small functional language consisting of:

\begin{itemize}
  \item Simple functional subset formed by lambda calculus with support for tuples and numbers.
  \item Comonadically-inspired primitives corresponding to \emph{counit}, \emph{cobind} and
    other operations of flat indexed comonads.
  \item Additional primitives that model contextual operations of each concrete coeffect language
    (the \kvd{prev} keyword, lookup function for the \ident{?p} syntax and operation that models
     implicit parameter binding for the $\kvd{let}~\ident{?p}=\ldots$ notation).
\end{itemize}
%
The syntax, typing and reduction rules of the first part (simple functional language) are common to
all concrete coeffect domain-specific languages. The syntax and typing rules of the second part
(comonadically-inspired primitives) are also shared by all coeffect DSLs, however the \emph{reduction
rules} for the comonadically-inspired primitives differ -- they capture the concrete notions of
context. Finally, the third part (domain-specific primitives) will differ for each coeffect
domain-specific language.

@@

The syntax of the target programming language is shown in Figure~\ref{fig:semantics-target}.
The values include numbers $n$, tuples and function values. The expressions include variables $x$,
values, lambda abstraction and application and operations on tuples. We do not need recursion or
other data types (although a realistic programming language would include them).
In what follows, we also use the following syntactic sugar for let binding:
%
\begin{equation*}
\begin{array}{rcl}
  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2)~e_1
\end{array}
\end{equation*}
%
Finally, $K[e]$ defines the syntactic evaluation context in which sub-expressions are evaluated.
Together with the evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the
standard call-by-value semantics of the common parts of the target language. The (standard) typing rules
for the common expressions of the target language are also shown in Figure~\ref{fig:semantics-target}.

@@

\begin{lemma}[Canonical forms]
\label{thm:semantics-fp-canon}
For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
\begin{enumerate}
  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
\end{enumerate}
\end{lemma}
\begin{proof}
  For (1), the last typing rule must have been (\emph{num}); for (2), it must have been
  (\emph{abs}) and for (3), the last typing rule must have been (\emph{tup}).
\end{proof}

@@

\begin{theorem}[Type preservation]
\label{thm:semantics-fp-pres}
  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$.
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow$.

@@

\begin{theorem}[Progress]
\label{thm:semantics-fp-prog}
  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$.
\end{theorem}
\begin{proof}
  By rule induction over $\vdash$.

@@

In this section, we use the same structure in a different way. Rather than treating the rules as
\emph{denotation} in categorical sense, we treat them as \emph{translation} from a source
domain-specific coeffect language into a target language with comonadically-inspired primitives
described in the previous section.

@@

\subsection{When is a coeffect not a monad}
\label{sec:semantics-related-monad}
